Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-06-07 figures.Rmd
added scatterplot of regressors
update for Friday seminar - convert coefficients to odds ratios

                "Deoxy Hb" = "cerebral.hbdeox", "Deoxy Hb α" = "cerebral.hbdeox.alpha", 
                "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
                "Cerebral O2 sat" = "cerebral.sat", "Cerebral O2 sat α" = "cerebral.sat.alpha", 
                "THC" = "cerebral.thc", "THCα" = "cerebral.thc.alpha")

@@

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
  dplyr::filter(`Deoxy Hb α` < 3)

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("Oxy Hb", "Oxy Hb α",   
                                                 "Deoxy Hb", "Deoxy Hb α", 
                                                "Cerebral O2 sat", "Cerebral O2 sat α", 
                                                "THC", "THCα")))

@@

df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC")) %>%

@@

# THC
kruskal.test(THC ~ Status, data = df_hbox_voi)
dunnTest(THC ~ Status, data = df_hbox_voi, method = "bh")

# Cerebral O2 sat
kruskal.test(`Cerebral O2 sat` ~ Status, data = df_hbox_voi)

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033). Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).

@@

df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%

@@

**Figure 2** Shown are the α values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).

@@

#### A. PCA
```{r fig3a}
# include variables of interest (THC, THC alpha, and hematocrit), remove healthy volunteers from analysis
df_hbox_log_voi <- df_hbox_impute %>% 
  dplyr::select(c(2,4:5,19:22)) %>% 
  dplyr::filter(Status != "HV")

@@

# select THC, THC alpha, and hematocrit
  df_hbox_model <- df_hbox_impute %>% 
    dplyr::select(c(2,4:5,19:22)) %>% 
    mutate(`1/10 * THCα` = THCα/10, `1/10 * THC` = THC/10) %>% 
    dplyr::select(-c(THCα, THC)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%

    dplyr::select(-c(`Cerebral O2 sat`, `Cerebral O2 sat α`, `Lactate (mmol/L)`))

@@

**Figure 3.** Logistic regression model discriminating CM vs UM patients using THC, THCα, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.

@@

hbox_final %>% 
  tidy() %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Supplementary figure 1: Spearman correlations of NIRS variables 
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

@@

m_cor_hv <- df_hbox_voi %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_voi %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_voi %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC", "Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%

@@

  # create plots
  m_cor_hv %>% f_plot_cor() + ggtitle("HV variable correlations") + theme(legend.position = "none")

  m_cor_um %>% f_plot_cor() + ggtitle("UM variable correlations") + theme(legend.position = "none")

  m_cor_cm %>% f_plot_cor() + ggtitle("CM variable correlations") + theme(legend.position = "none")

@@

### Supplementary figure 2: Logistic regression curves of NIRS variables
```{r fig_s2, fig.height = 12}

df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +

>>>>

                "Deoxy Hb" = "cerebral.hbdeox", "Deoxy Hb α" = "cerebral.hbdeox.alpha", 
                "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
                "HbO2%" = "cerebral.sat", "HbO2% α" = "cerebral.sat.alpha", 
                "[HbT]" = "cerebral.thc", "[HbT]α" = "cerebral.thc.alpha")

@@

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
  dplyr::filter(`Deoxy Hb α` < 3)

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("Oxy Hb", "Oxy Hb α",   
                                                 "Deoxy Hb", "Deoxy Hb α", 
                                                "HbO2%", "HbO2% α", 
                                                "[HbT]", "[HbT]α")))

@@

df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]")) %>%

@@

# [HbT]
kruskal.test(`[HbT]` ~ Status, data = df_hbox_voi)
dunnTest(`[HbT]` ~ Status, data = df_hbox_voi, method = "bh")

# HbO2%
kruskal.test(`HbO2%` ~ Status, data = df_hbox_voi)

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status [HbT] levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033). Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).

@@

df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  geom_hline(yintercept = 0.4, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 0.6, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +

  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +

  ylim(0, 1.5) +
  scale_color_manual(values = my_col) +
  theme_bw() +

  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15),
        legend.position = "none")

@@

```{r fig2.pvals, eval = FALSE}

# [HbT]α
kruskal.test(`[HbT]α` ~ Status, data = df_hbox_voi)
dunnTest(`[HbT]α` ~ Status, data = df_hbox_voi, method = "bh")

**Figure 2** Shown are the alpha values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).

@@

#### A. PCA
```{r fig3a}
# include variables of interest ([HbT], [HbT] alpha, and hematocrit), remove healthy volunteers from analysis
df_hbox_log_voi <- df_hbox_impute %>% 
  dplyr::select(c(2,4:5,19:22)) %>% 
  dplyr::filter(Status != "HV")

@@

# select [HbT], [HbT] alpha, and hematocrit
  df_hbox_model <- df_hbox_impute %>% 
    dplyr::select(c(2,4:5,19:22)) %>% 
    mutate(`1/10 * [HbT]α` = `[HbT]α`/10, `[HbT] (nM)` = `[HbT]`/1000) %>% 
    dplyr::select(-c(`[HbT]α`, `[HbT]`)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%

    dplyr::select(-c(`HbO2%`, `HbO2% α`, `Lactate (mmol/L)`))

@@

**Figure 3.** Logistic regression model discriminating CM vs UM patients using cerebral tissue hemoglobin concentration, cerebral tissue hemoglobin concentration alpha, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.

@@

hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Figure 4: Relationships among [HbT], [HbT] alpha, and hematocrit
```{r fig4}

df_hbox_impute %>% filter(Status != "HV") %>%

  ggplot(aes(x = `[HbT]`, y = `Hematocrit (%)`)) +
  geom_point(aes(color = Status, size = `[HbT]α`), alpha = 0.7) +

  xlab("[HbT] (μM)") +

**Figure 4.**

### Supplementary figure 1: Spearman correlations of potential regressors
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

@@

# create df of potential regressors
df_hbox_reg <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
  dplyr::filter(`Deoxy Hb α` < 3)

m_cor_hv <- df_hbox_reg %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_reg %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_reg %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]", "Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%

@@

  # create plots
  m_cor_hv %>% f_plot_cor() + ggtitle("Healthy volunteers") + theme(legend.position = "none")

  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria patients") + theme(legend.position = "none")

  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria patients") + theme(legend.position = "none")

@@

### Supplementary figure 2: Logistic regression curves of potential regressors
```{r fig_s2, fig.height = 12}

# pivot longer and change factor levels
df_hbox_reg_long <- df_hbox_reg %>% 
  pivot_longer(2:11, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Oxy Hb α", "Deoxy Hb", "Deoxy Hb α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))

# plot as individual regression curves
df_hbox_reg_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
